---
title: \textbf{\huge{Biochemical Process Calculation Tutorials}}
author: \textbf{\Large{Dhruva Punde}}
fontsize: 12pt
output: 
    pdf_document:
        extra_dependencies: ["fancyhdr", "amsmath", "hyperref"]
        toc: true
        toc_depth: 2
        number_sections: true
        df_print: kable
classoption: a4paper
---

\pagestyle{fancy}
\lhead{\large{Biochemical Process Calculation Tutorials}}
\rhead{}
\setlength{\headheight}{18.0pt}

\definecolor{shadecolor}{RGB}{230, 230, 230}

\newcommand{\chemeq}[1]{{\Large\[\mathrm{#1}\]}}
\newcommand{\ans}{\textbf{\underline{Answer}} $\rightarrow$}
\newcommand{\rulline}{\rule{456pt}{0.5mm}}

```{r echo=FALSE}
## 3 digits after decimal point
options(digits = 3)
```

\rulline

# Consider combustion of butane that forms the the stuff in LPG cylinder. Calculate for 1 kmol of butane burning:

## Kmol of O~2~ required.

## Kmol of air required.

## Kmol of CO~2~ produced.

## Kmol of H~2~O produced.

## Percent molar composition of product gas.

\ans

\chemeq{C_{4}H_{10} + 6.5O_{2} \rightarrow 4CO_{2} + 5H_{2}O \tag{1}}

\quad For the kmol of constituents produced and required, we have to multiply the coefficients by a known value, which in this case is that of butane (1 kmol). This calculation is done in the following program, along with the amount of air required and N~2~ present.

```{r, attr.source='.numberLines'}
## reactants and products
butane <- 1
oxygen <- butane*6.5 # part 1
carbondioxide <- butane*4 # part 3
water <- butane*5 # part 4

air <- oxygen*(100/21) # part 2
nitrogen <- oxygen*(79/21)
final <- carbondioxide + water + nitrogen # part 5
```

Kmol of O~2~ required is: **`r oxygen` kmol**  
Kmol of air required is: **`r air` kmol**  
Kmol of CO~2~ produced is: **`r carbondioxide` kmol**  
Kmol of H~2~O produced is: **`r water` kmol**  
Kmol of N~2~ required is: **`r nitrogen` kmol**  

Kmol of final gas produced is: **`r final` kmol**

The percentage composition is calculated by dividing the moles of the constituents by the sum of all the products times 100.

This function calculates the percentage composition by taking one parameter which is the constituent, multiplying it by 100 and dividing it by the sum of the products. It is used in the further numericals as well.

```{r, attr.source='.numberLines'}
percentage <- function(variable)
{
    (variable*100)/final
}
```

Percentage of CO~2~ is: **`r percentage(carbondioxide)`%**  
Percentage of H~2~O is: **`r percentage(water)`%**  
Percentage of N~2~ is: **`r percentage(nitrogen)`%**  

\rulline

# A coke is known to contain 90% carbon and 10% non-combustible ash (by weight).

## How many moles of O~2~ are theoretically required to burn 100 kg of coke completely.

## If 50% excess of air is supplied, calculate analysis of gases at end of combustion.

\ans

\chemeq{C + O_{2} \rightarrow CO_{2} \tag{1}}

\quad Since the coke contains 90% combustible carbon, we have to calculate the total amount of carbon present by weight and convert that into moles.  

The amount of O~2~ theoretically required and the amount of CO~2~ produced is then calculated by multiplying the coefficients by this value, which is calculated in the program below. The mass in kg is also calculated by multiplying the moles by the molecular weight of the constituents.

When 50% excess air is supplied, the unreacted O~2~ is also present in the products side, so the sum of all the products also includes this unreacted O~2~.

The percentage composition of the product gases is calculated by dividing the moles of the constituents by the sum of all the products times 100.

```{r, attr.source='.numberLines'}
## reactants and products
coke <- (90/100*100)/12 # 90% by weight to moles
oxygen <- coke*1 # part 1
carbondioxide <- coke*1

excess <- (oxygen*50/100) + oxygen # 50% excess, part 2
nitrogen <- excess*(79/21)
oxygen_at_end <- excess - oxygen # excess - theoretical
final <- nitrogen + oxygen_at_end + carbondioxide # product gas
```

Moles of O~2~ theoretically required to burn 100 kg of coke is: **`r oxygen` moles**, in kg that is: **`r oxygen*32` kg**  
Moles of O~2~ at end (excess) is: **`r oxygen_at_end` moles**, in kg that is: **`r oxygen_at_end*32` kg**  
Moles of N~2~ at the end is: **`r nitrogen` moles**, in kg that is: **`r nitrogen*28` kg**  
Moles of CO~2~ produced is: **`r carbondioxide` moles**, in kg that is: **`r carbondioxide*44` kg**

Percent of CO~2~ is: **`r percentage(carbondioxide)`%**  
Percent of O~2~ (excess) is: **`r percentage(oxygen_at_end)`%**  
Percent of N~2~ is: **`r percentage(nitrogen)`%**  

\rulline

# Find out quantity of reactant to produce 1000 kg of methanol.

\ans

\chemeq{CO + 2H_{2} \rightarrow CH_{3}OH \tag{1}}

\quad For 1000 kg of methanol, we first have to convert the weight into number of moles by dividing the weight by molecular weight.

The amount of CO and H~2~ required is then calculated by multiplying the coefficients by this value which is done in the program below. It is also converted to kg by multiplying the number of moles by the molecular weight.

```{r, attr.source='.numberLines'}
## reactants and products by moles
## dividing kg by molecular weight
methanol <- 1000/(12*1+4*1+16*1) 
carbonmono <- methanol*1
hydrogen <- methanol*2
```

Moles of CO required is: **`r carbonmono` moles**  
Moles of H~2~ required is: **`r hydrogen` moles**  

```{r, attr.source='.numberLines'}
## reactants and products by kg
## multiplying moles by molecular weight
methanol <- methanol*(12+1*4+16)
carbonmono <- carbonmono*(12+16)
hydrogen <- hydrogen*(2)
```

Amount of CO required is: **`r carbonmono` kg**  
Amount of H~2~ required is: **`r hydrogen` kg**  

\rulline

# A process in which 500 kg of CaCO~3~ produces CaO (calcination reaction) with 90% conversion. Prepare stoitiometric table.

\ans

\chemeq{CaCO_{3} \rightarrow CaO + CO_{2} \tag{1}}

\quad For 500 kg of CaCO~3~, we have to convert the weight to moles by dividing the weight by the molecular weight. Since 90% of CaCO~3~ is reacted, the moles of the products would also change accordingly, and the unreacted CaCO~3~ would be present in the product. The weight of the products can be calculated by multiplying the number of moles by the molecular weight.

In this program, there are a series of vectors which each correspond to weight, input moles, input mass, output moles and mass. A data frame is formed from these individual vectors and printed. The columns and rows and named accordingly. This forms the stoitiometric table.

```{r, attr.source='.numberLines'}
weight <- c((40+12+3*16), (40+16), (12+2*16)) # mol weights
i_moles <- c((500/weight[1]), 0, 0) # input moles
i_mass <- c((i_moles[1]*weight[1]), 0, 0) # input mass

caco3 <- (i_moles[1]*((100-90)/100))
cao <- (i_moles[1]-caco3)
carbondioxide <- (i_moles[1]-caco3)

o_moles <- c(caco3, cao, carbondioxide) # output moles
o_mass <- c((o_moles[1]*weight[1]), (o_moles[2]*weight[2]),
    (o_moles[3]*weight[3])) # output mass

## data frame to plot the stoitiometric table
## the previous vectors are used as rows and columns
table <- data.frame(weight, i_moles, i_mass, o_moles, o_mass)
colnames(table) <- c("weight", "input moles", 
    "input mass", "output moles", "output mass") # column names
rownames(table) <- c("CaCO3", "CaO", "CO2") # row names
table
```

\rulline

# Ethylene oxide is prepared by oxidation of ethylene. 100 kmol of ethylene and 100 kmol of O~2~ are charged to reactor. Percent conversion of ethylene is 85%, yield of ethylene oxide is 94.12%. Prepare stoitiometric table.

\ans

\quad First reaction:
\chemeq{C_{2}H_{4} + 0.5O_{2} \rightarrow C_{2}H_{4}O \tag{1}}

\quad Second reaction:
\chemeq{C_{2}H_{4} + 2O_{2} \rightarrow 2CO_{2} + 2H_{2}O \tag{2}}

100 kmol of O~2~ and ethylene are present. However the coefficient of O~2~ is $\frac{1}{2}$, thus ethylene is the limiting reagent. So only 50 kmol of O~2~ while 100 kmol of ethylene reacts.

The percent conversion of ethylene is 85%, so the number of moles of oxygen and ethylene oxide would be different than the coefficients. Also the yield of ethylene oxide is 94.12%, which further changes the number of moles of ethylene oxide produced. 

The remaining moles of ethylene is oxidised to CO~2~ and H~2~O. The moles of the CO~2~ and H~2~O produced can be calculated by multiplying the moles of the remaining ethylene by the coefficients of the products. So the exiting gas would contain ethylene oxide, CO~2~, H~2~O, unreacted ethylene and excess O~2~. This is calculated in the program below.

In this program, there are a series of vectors which correspond to weight, input moles, input mass, output moles, mass. A data frame is formed from these individual vectors and printed. The columns and rows and named accordingly. This forms the stoitiometric table.

Note: The mass of ethylene oxide does not add in the stoitiometric table due to the 94.12% yield.

```{r, attr.source='.numberLines'}
## reactants and products of the first reaction
ethylene <- (85/100)*100 # 85% conversion
oxygen <- ethylene/2
ethy_o <- ethylene*(94.12/100) # 94.12% yield
excess <- 100 # total oxygen supplied

## reactants and products of the second reaction
ethylene <- ethylene - ethy_o # remaining ethylene
carbondioxide <- ethylene*2
water <- ethylene*2
oxygen_at_end <- (excess-((oxygen)+(ethylene*2))) # unreacted oxygen

weight <- c((2*12+4*1), (2*16), (2*12+4*1+1*16), 
    (1*12+2*16), (2*1+1*16)) # molecular weights
i_moles <- c(100, 100, 0, 0, 0) # input moles
i_mass <- c((100*weight[1]), (100*weight[2]), 0, 0, 0) # input mass
o_moles <- c((100-(oxygen*2+ethylene)), oxygen_at_end, 
    ethy_o, carbondioxide, water) # output moles
o_mass <- c((weight[1]*o_moles[1]), (weight[2]*o_moles[2]),
    (weight[3]*o_moles[3]), (weight[4]*o_moles[4]),
        (weight[5]*o_moles[5])) # output mass

## data frame to plot the stoitiometric table
## the previous vectors are used as rows and columns
table <- data.frame(weight, i_moles, i_mass, o_moles, o_mass)
colnames(table) <- c("weight", "input moles", 
    "input mass", "output moles", "output mass") # column names
rownames(table) <- c("ethylene", "oxygen", "ethylene oxide", 
    "carbon dioxide", "water") # row names
table
```

\rulline

# Butane is burned in a combustion chamber.

## Calculate kmol of O~2~ required for 100 kmol of C~4~H~10~ fed for complete combustion.

## If we want to feed 20% excess over the theoretical value, calculate kmol of O~2~ to be fed for 100 kmol of butane.

## Calculate air fed to supply O~2~ calculated in 6.2 

## Calculate percentage molar composition of product gas assuming complete combustion with feed as in 6.3

\ans

\chemeq{C_{4}H_{10} + 6.5O_{2} \rightarrow 4CO_{2} + 5H_{2}O \tag{1}}

\quad For the kmol of constituents produced and required, we have to multiply the coefficients by a known value, which in this case is that of butane (100 kmol). The coefficients are then calculated in the following program.

For excess 20% O~2~, the number of moles of O~2~ would change, however since butane is the limiting reagent, the unreacted O~2~ would end up in the product side.

During incomplete combustion, CO is formed instead of CO~2~, this further changes the coefficients of the constituents.

When 80% of butane is reacted, the number of moles of butane changes, while changing the number of moles of the other coefficients as well. The unreacted butane is also present in the product side, which is also calculated by the program below.

```{r, attr.source='.numberLines'}
## reactants and products
butane <- 100
oxygen <- butane*6.5 # part 1
carbondioxide <- butane*4
water <- butane*5

excess <- (oxygen*(20/100)) + oxygen # 20% excess, part 2
air <- excess*(100/21) # part 3
nitrogen <- excess*(79/21)
oxygen_at_end <- excess - oxygen # excess - theoretical
final <- (carbondioxide + water + nitrogen
    + oxygen_at_end) # part 5
```

Kmol of O~2~ required for complete combustion is: **`r oxygen` Kmol**  
Kmol of 20% excess O~2~ fed to butane is: **`r excess` Kmol**  
Kmol of air fed is: **`r air` Kmol**  

Percentage of CO~2~ is: **`r percentage(carbondioxide)`%**  
Percentage of O~2~ (excess) is: **`r percentage(oxygen_at_end)`%**  
Percentage of H~2~O is: **`r percentage(water)`%**  
Percentage of N~2~ is: **`r percentage(nitrogen)`%**  

\rulline

# A combustion chamber is filled with butane and excess air. The composition of combustion gases (by volume) is CO~2~ = 9.39%, H~2~O = 11.73%, O~2~ = 4.70% and N~2~ = 74.13

## Find percent excess air used and mole ratio of air to butane used for complete combustion.

## For incomplete combustion, what are the elements of combustion gases.

## Find composition of combustion gases for 80% combustion of butane.

\ans

\chemeq{C_{4}H_{10} + 6.5O_{2} \rightarrow 4CO_{2} + 5H_{2}O \tag{1}}

\quad In this reaction, we assume 100 kmol of product gas is produced. Which gives us 9.39 kmol of CO~2~ and 11.73 kmol of H~2~O in the products. Dividing these values by the coefficients we can find out the number of moles of butane and O~2~, and then calculate the total amount of O~2~ by adding the unreacted oxygen in this value.

The composition of the product gas when 80% butane reacts is calculated by calculating the coefficient of butane and then
calculating the coefficients of the other constituents, this is done in the following program. The percentage composition of the product gas is calculated by dividing the number of moles of a constituent by the sum of the number of moles of the constituents in the product gas times 100.

```{r, attr.source='.numberLines'}
## reactants and products
carbondioxide <- 9.39
butane <- carbondioxide/4
oxygen <- butane*6.5
water <- butane*5

nitrogen <- 74.13
oxygen_at_end <- 4.70
excess <- oxygen + oxygen_at_end # end + theoretical
air <- excess*(100/21)
```

Mole ratio of air to butane for complete combustion is: **`r air` : `r butane`**

For incomplete combustion, the reaction is:

\chemeq{C_{4}H_{10} + 4.5O_{2} \rightarrow 4CO + 5H_{2}O \tag{2}}

For incomplete combustion, the elements of combustion gases are: **CO, H~2~O, O~2~ (excess) and N~2~**

For 80% combustion of butane,

```{r, attr.source='.numberLines'}
## reactants and products for 80% conversion
butane <- butane*(80/100)
oxygen <- butane*6.5
water <- butane*5
carbondioxide <- butane*4

oxygen_at_end <- excess - oxygen # excess - theoretical
final <- (water + carbondioxide + nitrogen
    + oxygen_at_end + (butane*(100/80)-butane)) # part 3
```

Percentage of CO~2~ is: **`r percentage(carbondioxide)`%**  
Percentage of H~2~O is: **`r percentage(water)`%**  
Percentage of N~2~ is: **`r percentage(nitrogen)`%**  
Percentage of O~2~ (excess) is: **`r percentage(oxygen_at_end)`%**  
Percentage of butane (unreacted) is: **`r percentage(butane)`%**  

\rulline

# Ethylene oxide is produced by oxidation of ethylene. 100 kmol of ethylene is fed to a reactor and product is found to contain 80 kmol of ethylene oxide and 10 kmol of CO~2~. Calculate percent conversion of ethylene and yield percentage of ethylene oxide.

\ans 

\quad First reaction:
\chemeq{C_{2}H_{4} + 0.5O_{2} \rightarrow C_{2}H_{4}O \tag{1}}

\quad Second reaction (after oxidation):
\chemeq{C_{2}H_{4} + 2O_{2} \rightarrow 2CO_{2} + 2H_{2}O \tag{2}}

Since 100 kmol of ethylene is fed to the reactor, theoretically 100 kmol of ethylene reacts with 50 kmol of O~2~ to produce 100 kmol of ethylene oxide. We also know that 10 kmol of CO~2~ is formed, from this we know that 5 kmol of ethylene reacts with 15 kmol of oxygen by looking at the coefficients.

```{r, attr.source='.numberLines'}
## second reaction
carbondioxide <- 10 
ethylene <- carbondioxide/2 

## first reaction
ethy_o <- 80
ethylene <- ethy_o + ethylene # ethylene oxide + ethylene
```

The percent conversion of ethylene can be calculated by dividing the moles of ethylene reacted to produce Ethylene Oxide by the number of moles present times 100.

The yield percentage can be calculated by dividing the moles of ethylene oxide produced by the number of moles of ethylene present in the first reaction times 100.

Percent conversion of Ethylene Oxide is: **`r ethy_o/100*100`%**  
Percent yield of Ethylene Oxide is: **`r ethy_o/ethylene*100`%**  

\rulline

# A gas analysing CO~2~ = 5.5%, CO = 25%, H~2~ = 14% CH~4~ = 0.5% and N~2~ = 55% (by pressure) is burnt in furnace with air which is 10% excess over that required to burn CO, H~2~, CH~4~ completely. Give the analysis of product gas reaction assuming all reaction to completion.

\ans

\quad Reactions:

\chemeq{2CO + O_{2} \rightarrow 2CO_{2} \tag{1}}

\chemeq{2H_{2} + O_{2} \rightarrow 2H_{2}O \tag{2}}

\chemeq{CH_{4} + 2O_{2} \rightarrow CO_{2} + H_{2}O \tag{3}}

The percentage of gases in the furnace are by pressure, from the $PV = nRT$, Ideal Gas Law, we know that pressure percentage = volume percentage = molar percentage. So we shall assume the percentages as moles per 100 moles. However in the reactions above, we see that products like CO~2~ and H~2~O are also formed, these products will also be present in the furnace. We can find out the number of moles of these products by looking at the coefficients of the reactants.

From the above reactions, we can also calculate the theoretical number of moles of O~2~ required to burn CO, H~2~ and CH~4~, which would be adding the coefficients of oxygen for all the 3 reactions, which is done by the program below.

```{r, attr.source='.numberLines'}
## moles of gases present in furnace
carbondioxide <- (5.5) + (25/2*2) + (0.5*1) # furnace + reaction 1 and 3
carbonmono <- 25
hydrogen <- 14
methane <- 0.5
nitrogen <- 55
water <- (14/2*2) + (0.5*1) # reaction 2 and 3

## amount of theoretical and excess oxygen
oxygen <- ((carbonmono/2) + (hydrogen/2)
    + (methane*2)) # theoretical oxygen
excess <- (oxygen*10/100)+oxygen # 10% excess
oxygen_at_end <- excess - oxygen # excess - theoretical

## moles of nitrogen, furnace + moles in air
nitrogen <- nitrogen + (79/21*excess)

## total moles of exiting gas
final <- nitrogen + oxygen_at_end + carbondioxide + water
```

The gas at the end would contain N~2~, excess O~2~, CO~2~ and H~2~O. The analysis of gases can be calculated by dividing the moles of a particular gas by the summation of the total moles times 100.

Percentage of N~2~ is: **`r percentage(nitrogen)`%**  
Percentage of O~2~ (excess) is: **`r percentage(oxygen_at_end)`%**  
Percentage of CO~2~ is: **`r percentage(carbondioxide)`%**  
Percentage of H~2~O is: **`r percentage(water)`%**  

\rulline

# A mixture of pure CO~2~ and H~2~ is passed over a Nickel catalyst. The temperature of catalyst bed is 588K and reacting pressure is 2.02 Mega Pascal gauge. The gas mixture leaving the reactor is analysed to contain 57.1% CO~2~, 41.1% of H~2~, 1.68% of CH~4~ and 0.12% of CO by volume.

## Calculate percent conversion of CO~2~.

## Calculate yield of methane in terms of CO~2~ reacted.

## Calculate composition of feed.

\ans 

\quad Reactions: 

\chemeq{CO_{2} + 4H_{2} \rightarrow CH_{4} + 2H_{2}O \tag{1}}

\chemeq{CO_{2} + H_{2} \rightarrow CO + H_{2}O \tag{2}}

Since the percentages are by volume, we can assume them to be mole percentages. The exiting gas contains CO~2~ and H~2~, so we can assume that CO~2~ and H~2~ do not react completely in the reactor and are present in excess.

The exiting gas contains CH~4~ and CO, which are the products of the reactions above. Through this we can calculate the amount of CO~2~ and H~2~ theoretically required to react which produces the products by multiplying the moles of the products by the coefficients of the reactants.

```{r, attr.source='.numberLines'}
## products from the reactions 
methane <- 1.68
carbonmono <- 0.12

## theoretical amount of hydrogen and carbondioxide
carbondioxide <- methane + carbonmono
hydrogen <- (methane*4) + carbonmono

## amount of carbondioxide and hydrogen fed in the reactor
carbondioxide <- carbondioxide + 57.1
hydrogen <- hydrogen + 41.1
```

Percent conversion of CO~2~ is calculated by dividing the moles of CO~2~ reacted by the total CO~2~ fed times 100.

Yield of CH~4~ is calculated by dividing the number of moles of CH~4~ by the total moles of CO~2~ times 100. 

Percent conversion of CO~2~ is: **`r (carbondioxide-57.1)/carbondioxide*100`%**  
Yield of methane in terms of CO~2~ is: **`r methane/carbondioxide*100`%**  
The composition of feed is **`r carbondioxide` moles of CO~2~** and **`r hydrogen` moles of H~2~**  

\rulline

\begin{flushleft}
    \footnotesize
    \textit{source code of the document:}
    \url{https://github.com/dhruvapunde/biochemical-programs/blob/main/bcp-tutorials-sem-3.Rmd}
\end{flushleft}
