---
title: Biochemical Process Calculation Tutorials
author: Dhruva Punde 
output: 
    pdf_document:
        extra_dependencies: ["amsmath", "amssymb", "blindtext"]
        toc: true
        toc_depth: 2
        number_sections: true
        df_print: kable
---

\newcommand{\chemeq}[1]{$$\mathrm{#1}$$}
\newcommand{\ans}[1]{\textbf{\underline{#1}}}

\Large

# Consider combustion of butane that forms the the stuff in LPG cylinder. Calculate for 1 kmol of butane burning:

## Kmol of O~2~ required

## Kmol of air required

## Kmol of CO~2~ produced

## Kmol of H~2~O produced

## Percent molar composition of flue gas

\ans{Answer} $\rightarrow$

\chemeq{C_{4}H_{10} + 6.5O_{2} \rightarrow 4CO_{2} + 5H_{2}O}

```{r echo=FALSE}
## 3 digits afer decimal point
options(digits=3)
```

```{r}
## reactants and products
butane <- 1
oxygen <- 6.5 # part 1
carbondioxide <- 4 # part 3
water <- 5 # part 4

air <- 100/21*oxygen # part 2
nitrogen <- 79/21*oxygen
final <- carbondioxide + water + nitrogen # part 5
```

Kmol of O~2~ required is: **`r oxygen` kmoles**  
Kmol of air required is: **`r air` kmoles**  
Kmol of CO~2~ produced is: **`r carbondioxide` kmoles**  
Kmol of H~2~O produced is: **`r water` kmoles**  
Kmol of N~2~ required is: **`r nitrogen` kmoles**  

Kmol of final gas produced is: **`r final` kmoles**

Percentage of CO~2~ is: **`r carbondioxide*100/final`%**  
Percentage of H~2~O is: **`r water*100/final`%**  
Percentage of N~2~ is: **`r nitrogen*100/final`%**  

# A coke is known to contain 90% carbon and 10% non-combustible ash (by weight).

## How many moles of O~2~ are theoretically required to burn 100 kg of coke completely.

## If 50% excess of air is supplied, calculate analysis of gases at end of combustion

\ans{Answer} $\rightarrow$

\chemeq{C + O_{2} \rightarrow CO_{2}}

```{r}
coke <- (90/100*100)/12 # 90% by weight to moles
oxygen <- coke*1
carbondioxide <- coke*1

nitrogen <- 79/21*oxygen
excess <- (50/100*oxygen)+oxygen # 50% excess
oxygen_at_end <- excess - oxygen # excess - theoretical
final <- nitrogen + oxygen_at_end + carbondioxide
```

Moles of O~2~ theoretically required to burn 100 kg of coke is: **`r oxygen` moles**, in kg that is: **`r oxygen*32` kg**  
Moles of O~2~ at end (excess) is: **`r oxygen_at_end` moles**, in kg that is: **`r oxygen_at_end*32` kg**  
Moles of N~2~ at the end is: **`r nitrogen` moles**, in kg that is: **`r nitrogen*28` kg**  
Moles of CO~2~ produced is: **`r carbondioxide` moles**, in kg that is: **`r carbondioxide*44` kg**

Percent of CO~2~ is: **`r carbondioxide*100/final`%**  
Percent of O~2~ (excess) is: **`r oxygen_at_end*100/final`%**  
Percent of N~2~ is: **`r nitrogen*100/final`%**  

# Find out quantity of reactant to produce 1000 kg of methanol.

\ans{Answer} $\rightarrow$

\chemeq{CO + 2H_{2} \rightarrow CH_{3}OH}

```{r}
methanol <- 1000/32 # kg to moles
carbonmono <- 1*methanol
hydrogen <- 2*methanol

carbonmono <- carbonmono*28
hydrogen <- hydrogen*2
```

CO required is: **`r carbonmono` kg**, that in Kmol is: **`r carbonmono/28` Kmol**  
H~2~ required is: **`r hydrogen` kg**, that in Kmol is: **`r hydrogen/2` Kmol**  

# A process in which 500 kg of CaCO~3~ produces CaO (calcination reaction) with 90%. Prepare stoitiometric table.

\ans{Answer} $\rightarrow$

\chemeq{CaCO_{3} \rightarrow CaO + CO_{2}}

```{r}
moles <- 500/100 # kg to moles
p_moles <- 90/100*moles
weight <- c((40+12+3*16), (40+16), (12+2*16))
i_moles <- c(moles, 0, 0)
i_mass <- c(500, 0, 0)
moles <- moles - p_moles
o_moles <- c(moles, p_moles, p_moles)
mass <- c(500, 4.5*(40+16), 4.5*(12+2*16))

## data frame to plot the stoitiometric table
table <- data.frame(
weight, i_moles, i_mass, o_moles, mass)
colnames(table) <- c("weight", "input moles", 
"input mass", "output moles", "mass")
rownames(table) <- c("CaCO3", "CaO", "CO2")
table
```

# Ethylene oxide is prepared by oxidation of ethylene. 100 kmol of ethylene and 100 kmol of O~2~ are charged to reactor. Percent conversion of ethylene is 85%, yield of ethylene oxide is 94.12%. Prepare stoitiometric table.

\ans{Answer} $\rightarrow$


















# A combustion chamber is filled with butane and excess air. The composition of combustion gases (by volume) is CO~2~ = 9.39%, H~2~O = 11.73%, O~2~ = 4.70% and N~2~ = 74.13

## Find percent excess air used and mole ratio of air to butane used for complete combustion.

## For incomplete combustion, what are the elements of combustion gases.

## Find composition of combustion gases for 80% combustion of butane.

## Find percent excess air and percent conversion of butane and quantity of butane and air used, if combustion of gases on pressure basis is given by CO~2~ = 8%, H~2~O = 11.2%, butane = 1.8%, O~2~ = 4.8% and N~2~ = 74.2%

\ans{Answer} $\rightarrow$

\chemeq{C_{4}H_{10} + 6.5O_{2} \rightarrow 4CO_{2} + 5H_{2}O}













